---
title: "ANSWER KEY — Exploring Human Health Data with ggplot2 and dplyr"
author: "Instructor: David Gresham"
date: "`r Sys.Date()`"
output: 
  html_notebook: 
    toc: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Part 1 — Loading the data

**1.1**
```{r load-data}
library(tidyverse)

health <- read_tsv("human_health_data.tsv")
head(health)
```

**1.2**
```{r glimpse}
glimpse(health)
```

---

## Part 2 — Selecting and filtering

**2.1**
```{r select}
clinical <- health |>
  select(sample_id, sex, blood_glucose_mgdl, LDL_mgdl)

clinical
```

**2.2**
```{r filter}
prediabetic_female <- health |>
  filter(sex == "Female", blood_glucose_mgdl > 100)

nrow(prediabetic_female)
```

**2.3**
```{r filter2}
brown_or_blue <- health |>
  filter(hair_color == "Brown" | eye_color == "Blue")

nrow(brown_or_blue)
```

---

## Part 3 — Creating new variables

**3.1**
```{r mutate-bmi}
health <- health |>
  mutate(bmi = round(weight_kg / (height_cm / 100)^2, 2))

health |> select(sample_id, weight_kg, height_cm, bmi) |> head()
```

**3.2**
```{r mutate-glucose}
health <- health |>
  mutate(glucose_category = case_when(
    blood_glucose_mgdl < 100  ~ "Normal",
    blood_glucose_mgdl <= 125 ~ "Pre-diabetic",
    blood_glucose_mgdl > 125  ~ "Diabetic"
  ))

health |> count(glucose_category)
```

---

## Part 4 — Summarizing by group

**4.1**
```{r summarize-bmi}
health |>
  group_by(sex) |>
  summarize(
    mean_bmi = mean(bmi),
    sd_bmi   = sd(bmi)
  )
```

**4.2**
```{r count-glucose}
health |>
  group_by(sex, glucose_category) |>
  summarize(n = n(), .groups = "drop")
```

**4.3**
```{r summarize-ldl}
health |>
  group_by(hair_color) |>
  summarize(mean_LDL = mean(LDL_mgdl)) |>
  arrange(desc(mean_LDL))
```

---

## Part 5 — Visualization with ggplot2

**5.1 Histogram**
```{r hist-glucose}
ggplot(health, aes(x = blood_glucose_mgdl, fill = sex)) +
  geom_histogram(binwidth = 5, alpha = 0.6, position = "identity") +
  labs(
    title = "Distribution of Fasting Blood Glucose",
    x     = "Blood Glucose (mg/dL)",
    y     = "Count",
    fill  = "Sex"
  )
```

**5.2 Scatter — height vs. weight**
```{r scatter-hw}
ggplot(health, aes(x = height_cm, y = weight_kg, color = sex)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Height vs. Weight by Sex",
    x     = "Height (cm)",
    y     = "Weight (kg)",
    color = "Sex"
  )
```

**5.3 Box plot — BMI by sex**
```{r boxplot-bmi}
ggplot(health, aes(x = sex, y = bmi, fill = sex)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4) +
  labs(
    title = "BMI Distribution by Sex",
    x     = "Sex",
    y     = "BMI (kg/m²)"
  ) +
  theme(legend.position = "none")
```

**5.4 Bar chart — hair color by sex**
```{r bar-hair}
ggplot(health, aes(x = hair_color, fill = sex)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Sample Counts by Hair Color and Sex",
    x     = "Hair Color",
    y     = "Count",
    fill  = "Sex"
  )
```

**5.5 Scatter — BMI vs. LDL**
```{r scatter-bmi-ldl}
ggplot(health, aes(x = bmi, y = LDL_mgdl, color = glucose_category)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title  = "BMI vs. LDL Cholesterol by Glucose Category",
    x      = "BMI (kg/m²)",
    y      = "LDL (mg/dL)",
    color  = "Glucose Category"
  )
```

**5.6 Faceted plot**
```{r facet-glucose-ldl}
ggplot(health, aes(x = blood_glucose_mgdl, y = LDL_mgdl)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", color = "firebrick") +
  facet_wrap(~ sex) +
  labs(
    title = "Blood Glucose vs. LDL by Sex",
    x     = "Blood Glucose (mg/dL)",
    y     = "LDL (mg/dL)"
  )
```

---

## Part 6 — Open-ended example

One example: violin plot of LDL by glucose category, filled by sex.

```{r open-ended}
health |>
  mutate(glucose_category = factor(glucose_category,
                                   levels = c("Normal", "Pre-diabetic", "Diabetic"))) |>
  ggplot(aes(x = glucose_category, y = LDL_mgdl, fill = sex)) +
  geom_violin(alpha = 0.6, position = position_dodge(0.8)) +
  geom_boxplot(width = 0.1, position = position_dodge(0.8), outlier.shape = NA) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "LDL Distribution by Glucose Category and Sex",
    x     = "Glucose Category",
    y     = "LDL (mg/dL)",
    fill  = "Sex"
  )
```

*Observations:* Samples in the diabetic and pre-diabetic categories tend to
have higher LDL values and greater spread, particularly among males. This
aligns with the known co-occurrence of dyslipidemia and impaired glucose
metabolism. The normal-glucose group shows a tighter LDL distribution in
both sexes.
