---
title: "Exploring Human Health Data with ggplot2 and dplyr"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

In this exercise you will work with a dataset of 200 human samples containing
physical measurements and clinical lab values. You will practice:

- Reading tabular data into R
- Manipulating data with **dplyr** (`filter`, `select`, `mutate`, `group_by`, `summarize`)
- Visualizing data with **ggplot2**

The dataset (`human_health_data.tsv`) contains the following columns:

| Column | Description |
|--------|-------------|
| `sample_id` | Unique sample identifier |
| `sex` | Biological sex (Male / Female) |
| `height_cm` | Height in centimetres |
| `weight_kg` | Weight in kilograms |
| `hair_color` | Hair color |
| `eye_color` | Eye color |
| `blood_glucose_mgdl` | Fasting blood glucose (mg/dL) |
| `LDL_mgdl` | LDL cholesterol (mg/dL) |

---

## Part 1 — Loading the data

**1.1** Load the `tidyverse` package and read `human_health_data.tsv` into a
tibble called `health`. Use `read_tsv()`. Print the first few rows with
`head()`.

```{r load-data}
library(tidyverse)

#Read in health data
data <- read_tsv("human_health_data.tsv")



```

**1.2** Use `glimpse()` to get an overview of the data types in `health`.

```{r glimpse}
# Your code here
glimpse(data)
```

---

## Part 2 — Selecting and filtering (`select`, `filter`)

**2.1** Create a new tibble called `clinical` that contains only the columns
`sample_id`, `sex`, `blood_glucose_mgdl`, and `LDL_mgdl`.

```{r select}
#Create a new tibble called `clinical` that contains only the columns `sample_id`, `sex`, `blood_glucose_mgdl`, and `LDL_mgdl`.
mytiblle <- data %>%
  select(sample_id,
          sex,
          blood_glucose_mgdl,
          LDL_mgdl)


```

**2.2** From the original `health` tibble, filter to keep only **Female**
samples whose `blood_glucose_mgdl` is greater than 100 mg/dL (pre-diabetic
range). How many samples meet this criterion?

```{r filter}
# 

```

**2.3** Filter `health` to keep only samples with brown hair **or** blue eyes.
Store the result as `brown_or_blue`.

```{r filter2}
# Your code here

```

---

## Part 3 — Creating new variables (`mutate`)

**3.1** Add a column called `bmi` (Body Mass Index) to `health` using the
formula:

$$BMI = \frac{weight\_kg}{(height\_cm / 100)^2}$$

Round to two decimal places. Save the result back to `health`.

```{r mutate-bmi}
# Your code here

```

**3.2** Add a column called `glucose_category` that classifies each sample:

- `"Normal"` if `blood_glucose_mgdl < 100`
- `"Pre-diabetic"` if between 100 and 125 (inclusive)
- `"Diabetic"` if `blood_glucose_mgdl > 125`

*Hint:* Use `case_when()` inside `mutate()`.

```{r mutate-glucose}
# Your code here

```

---

## Part 4 — Summarizing by group (`group_by`, `summarize`)

**4.1** Calculate the mean and standard deviation of `bmi` separately for
Males and Females. Your output should have three columns: `sex`, `mean_bmi`,
and `sd_bmi`.

```{r summarize-bmi}
# Your code here

```

**4.2** Count how many samples fall into each `glucose_category`, broken down
by `sex`. Your output should have columns `sex`, `glucose_category`, and `n`.

```{r count-glucose}
# Your code here

```

**4.3** For each `hair_color`, calculate the mean `LDL_mgdl`. Arrange the
result from highest to lowest mean LDL.

```{r summarize-ldl}
# Your code here

```

---

## Part 5 — Visualization with ggplot2

For each plot, make sure to add informative axis labels and a title using
`labs()`.

**5.1 Histogram — distribution of blood glucose**

Plot a histogram of `blood_glucose_mgdl`. Color the bars by `sex` using
`fill`. Use `alpha = 0.6` so overlapping bars are visible. Set appropriate
binwidth.

```{r hist-glucose}
# Your code here

```

**5.2 Scatter plot — height vs. weight**

Create a scatter plot of `height_cm` (x-axis) vs. `weight_kg` (y-axis),
coloring points by `sex`. Add a linear regression line for each sex using
`geom_smooth(method = "lm")`.

```{r scatter-hw}
# Your code here

```

**5.3 Box plot — BMI by sex**

Make a box plot of `bmi` for each `sex`. Overlay the individual data points
using `geom_jitter()` with `width = 0.2` and `alpha = 0.4`.

```{r boxplot-bmi}
# Your code here

```

**5.4 Bar chart — sample counts by hair color and sex**

Using `geom_bar()`, create a bar chart showing the number of samples for each
`hair_color`, filled by `sex`. Use `position = "dodge"` so the bars for each
sex are side by side.

```{r bar-hair}
# Your code here

```

**5.5 Scatter plot — BMI vs. LDL**

Plot `bmi` (x) vs. `LDL_mgdl` (y), coloring points by `glucose_category`.
Choose a colorblind-friendly palette with `scale_color_brewer(palette = "Set2")`.
Does there appear to be a relationship?

```{r scatter-bmi-ldl}
# Your code here

```

**5.6 Faceted plot — glucose vs. LDL by sex**

Plot `blood_glucose_mgdl` (x) vs. `LDL_mgdl` (y) as a scatter plot, then use
`facet_wrap(~ sex)` to create separate panels for Males and Females.

```{r facet-glucose-ldl}
# Your code here

```

---

## Part 6 — Putting it together (open-ended)

**6.1** Using any combination of `dplyr` and `ggplot2`, create one additional
plot of your choice that reveals something interesting about this dataset.
Write 2–3 sentences below your plot describing what you observe.

```{r open-ended}
# Your code here

```

*Your observations:*

---

## Submission

Knit this document to HTML and submit the `.html` file.
